Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-05 paper/paper.tex
additional clean up

The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and/or orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the site-site interactions could be orientationally-dependent (e.g., point dipoles).
The orientation of a particle might be given implicitly (not part of $\boldsymbol{\omega}$) via the Cartesian coordinates of its interaction sites.
Alternatively, it may be specified explicitly via an appropriate orientation vector, with the location of the interaction sites (if a multi-site particle) constructed on-the-fly from the position, orientation, and internal coordinates.
In a 3-dimensional space, the orientation coordinate for a cylindrically-symmetric object requires just two independent values, e.g., a unit  vector describing the cylindrical axis direction.
For particles with no such symmetry, three independent values are needed to specify the orientation.
Euler angles ($\phi\theta\psi$) are sometimes invoked for this, but they are not particularly useful in the context of molecular simulation.
A better choice is a quaternion ${\bf q}$, which is a normalized, four-element vector \cite{vesely_angular_1982, karney_quaternions_2007}.

@@

Translation contains only the linear part of a displacement and not angular displacement.
An MC translation trial move is defined as follows.
A particle is randomly chosen among all of the mobile, non-fixed particles in the old microstate.
The total number of mobile particles is $N_m=\sum_i^{mobile} N_i$, where $i$ is the particle type.
The randomly chosen particle with position $\mathbf{r}_o$ in the old microstate is then translated by a random amount selected uniformly within $\pm\delta/2$ in each of $D$ dimensions, to a new position labeled $\mathbf{r}_n$, thereby defining the new microstate.

@@

Substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance} and using the transition probabilities in Table~\ref{tab:lhs_rotation} yields,
\begin{equation}
\chi = e^{-\beta\Delta U}
\label{eq:lhs_rotate}
\end{equation}
for the Metropolis acceptance of particle orientation changes.
%For a rigid rotation of a particle in three dimensions, the orientation may be given by Euler angles or quaternions.
%However, when obtaining a random orientation, it is important not to select random values uniformly within the range of each of these variables.
%For example, independent and uniform selection of Euler angles will bias orientations toward the poles and makes the calculation of $P_{\omega n}$ and $P_{\omega o}$ nontrivial.
%Instead, the orientation should be uniformly chosen on the surface of a sphere \cite{allen_computer_1989}.
%In that case, $P_{\omega n}=P_{\omega o}$.
%The d$\boldsymbol{\omega}$ terms in the probability of states, $\Pi$, include the orientations of the particles, which are sampled uniformly over the entire surface of a sphere and are normalized with the appropriate Jacobian.
%An example of randomly picking points on the surface of a sphere is shown in Section~\ref{sec:unit_sphere}.
%
%A more efficient approach for sampling particle orientations is to perturb the rigid orientation using a tunable parameter, $\delta$.
%This must also be done in a careful way and not simply by changing the Euler angles uniformly and independently.
%One example algorithm to perturb orientation proceeds as follows.
%Generate a random quaternion, $\mathbf{q}$, uniformly on the surface of a sphere \cite{vesely_angular_1982}.
%If the original particle orientation is represented by a quaternion, $\mathbf{q}_o$, then the newly proposed orientation is given by the quaternion $\mathbf{q}_n=\frac{\mathbf{q}_o+\delta\mathbf{q}}{|\mathbf{q}_o+\delta\mathbf{q}|}$.
%If the original particle orientation is represented by a collection of coordinates, use $\mathbf{q}_o=[0, 0, 0, 1]$ (which corresponds to an identity rotation matrix) to generate a rotation matrix from $\mathbf{q}_n$ to perturb the coordinates.
%Because of the use of the tunable parameter, $P_{\omega n}$ and $P_{\omega o}$ are equal to each other and therefore cancel one another in Eq.~\ref{eq:lhs_rotate}.

@@

%A similar problem is sampling a quaternion, which may be interpreted as a point on the surface of a unit 4D-hypersphere. Accordingly, Code Block~\ref{sim:quaternion} shows how quaternions may be randomly generated by sampling uniformly on the surface of a 4D hypersphere using an efficient rejection method \cite{vesely_angular_1982}.
%Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices, which could be used to generate a fully random orientation by rotating from an arbitrary reference orientation.
Finally, a randomly generated perturbation of orientation is shown using the axis-angle method in Fig.~\ref{fig:random_orientation}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
  Example output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random unit vectors emanating from the origin and $10^3$ unit vectors in orange randomly generated about the upwards $z$-axis (shown in black) with axis-angle rotations of a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees (e.g., the orange vectors are confined to the top hemisphere).
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/spherical_shell.png}
\caption{
Example output from Code Block~\ref{sim:spherical_shell} shows approximately $10^5$ small blue points randomly generated inside a spherical shell of outer radius $3$ and inner radius $2$ and cut in half at $z=0$.
\label{fig:spherical_shell}
}
\end{centering}
\end{figure}

>>>>

The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and/or orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the site-site interactions could be orientationally-dependent (e.g., point dipoles).
The orientation of a particle might be given implicitly (not part of $\boldsymbol{\omega}$) via the Cartesian coordinates of its interaction sites.
Alternatively, it may be specified explicitly via an appropriate orientation vector, with the location of the interaction sites (if a multi-site particle) constructed on-the-fly from the position, orientation, and internal coordinates.
In a 3-dimensional space, the orientation coordinate for a cylindrically-symmetric object requires just two independent values, e.g., a unit  vector describing the cylindrical axis direction.
For particles with no such symmetry, three independent values are needed to specify the orientation.
Euler angles are sometimes invoked for this, but they are not particularly useful in the context of molecular simulation.
A better choice is a quaternion, which is a normalized, four-element vector \cite{vesely_angular_1982, karney_quaternions_2007}.

@@

An MC translation trial move is defined as follows.
A particle is randomly chosen among all of the mobile, non-fixed particles in the old microstate.
The total number of mobile particles is $N_m=\sum_i^{mobile} N_i$, where $i$ is the particle type.
The randomly chosen particle with position $\mathbf{r}_o$ in the old microstate is then translated by a random amount selected uniformly within $\pm\delta/2$ in each of $D$ dimensions, to a new position labeled $\mathbf{r}_n$, thereby defining the new microstate.

@@

Substituting Eq.~\ref{eq:rhs_nvt} into Eq.~\ref{eq:detailed_balance} and using the transition probabilities in Table~\ref{tab:lhs_rotation} yields,
\begin{equation}
\chi = e^{-\beta\Delta U}
\label{eq:lhs_rotate}
\end{equation}
for the Metropolis acceptance of particle orientation changes.
To simplify notation, all remaining trial moves which include changes in orientation will simply refer to these transition probabilities as $P_{\omega n}$ and $P_{\omega o}$ for the new and old states, respectively.

%For a rigid rotation of a particle in three dimensions, the orientation may be given by Euler angles or quaternions.
%However, when obtaining a random orientation, it is important not to select random values uniformly within the range of each of these variables.
%For example, independent and uniform selection of Euler angles will bias orientations toward the poles and makes the calculation of $P_{\omega n}$ and $P_{\omega o}$ nontrivial.
%Instead, the orientation should be uniformly chosen on the surface of a sphere \cite{allen_computer_1989}.
%In that case, $P_{\omega n}=P_{\omega o}$.
%The d$\boldsymbol{\omega}$ terms in the probability of states, $\Pi$, include the orientations of the particles, which are sampled uniformly over the entire surface of a sphere and are normalized with the appropriate Jacobian.
%An example of randomly picking points on the surface of a sphere is shown in Section~\ref{sec:unit_sphere}.
%
%A more efficient approach for sampling particle orientations is to perturb the rigid orientation using a tunable parameter, $\delta$.
%This must also be done in a careful way and not simply by changing the Euler angles uniformly and independently.
%One example algorithm to perturb orientation proceeds as follows.
%Generate a random quaternion, $\mathbf{q}$, uniformly on the surface of a sphere \cite{vesely_angular_1982}.
%If the original particle orientation is represented by a quaternion, $\mathbf{q}_o$, then the newly proposed orientation is given by the quaternion $\mathbf{q}_n=\frac{\mathbf{q}_o+\delta\mathbf{q}}{|\mathbf{q}_o+\delta\mathbf{q}|}$.
%If the original particle orientation is represented by a collection of coordinates, use $\mathbf{q}_o=[0, 0, 0, 1]$ (which corresponds to an identity rotation matrix) to generate a rotation matrix from $\mathbf{q}_n$ to perturb the coordinates.
%Because of the use of the tunable parameter, $P_{\omega n}$ and $P_{\omega o}$ are equal to each other and therefore cancel one another in Eq.~\ref{eq:lhs_rotate}.

@@

Code Block~\ref{sim:rotationmatrix} implements Eq.~\ref{eq:rotation_matrix} which are used to generate orientation-change trials in Code Block~\ref{sim:random_orientation} with the axis-angle method shown in Fig.~\ref{fig:random_orientation}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
The output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random blue unit vectors emanating from the origin.
The other $10^3$ orange unit vectors are confined to the top hemisphere because they were randomly generated with the axis-angle method by perturbing the orientation of an upwards $z$-axis vector (shown in black) with a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees.
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/spherical_shell.png}
\caption{
Example output from Code Block~\ref{sim:spherical_shell} shows approximately $10^5$ small blue points randomly generated inside a spherical shell of outer radius $3$ and inner radius $2$.
To see the inside cavity, points above $z=0$ were removed.
\label{fig:spherical_shell}
}
\end{centering}
\end{figure}
